---
title: "Education quality"
author: "Daphné PERRY Justine CATTA Chloé MARQUIS Clémence HUANG"
execute:
  echo: false
format:
  wordcount-html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
  pdf:
    toc: true
editor: visual
---

1.  

    ```         
    What are the different high school quality indicators published by the ministry of higher education? [^sources1]
    ```
[^sources1]:
  https://www.education.gouv.fr/les-indicateurs-de-resultats-des-colleges-et-des-lycees-377729
  https://data.education.gouv.fr/explore/dataset/fr-en-indicateurs-de-resultat-des-lycees-gt_v2/ 


The French Ministry of Education, through the DEPP (Direction de l’évaluation, de la prospective et de la performance), publishes annually the *High School Value-Added Indicators* (Indicateurs de Valeur Ajoutée des Lycées, IVAL).\
Their objective is to assess high school performance while accounting for differences in student intake - rather than relying solely on raw outcome measures.

Three main **quality indicators** are published for general and technological high schools:

\underline{Baccalaureate success rate}

This indicator measures the proportion of students present at the baccalaureate examination who obtain the diploma: \[ \text{Success rate} = \frac{\text{Number of students who pass the bac}}{\text{Number of students present at the exam}} \]

\underline{Access rate to the baccalaureate}

The access rate represents the probability that a student entering a given high school in *seconde* will eventually obtain the baccalaureate in the same establishment, even if the student repeats a grade.\
It is computed as the product of successive transition probabilities: \[ \text{Access rate}\_{2nde \to bac} = (2nde \to 1ère) \times (1ère \to Terminale) \times (Terminale \to Bac) \]

\underline{Baccalaureate honors (mentions) rate}

This indicator measures the proportion of students present at the baccalaureate examination who obtain a mention (Assez Bien, Bien, or Très Bien (with or without Félicitations)).

For most indicators, results are published both **overall** and **by track or series** (general and technological streams such as STI2D, STMG, STL, ST2S, etc.).\
Each indicator is reported as an **observed value**, an **expected value**, and a **value-added measure**.

Administrative variables available in the datasets (UAI code, location, sector, student counts by grade) are descriptive and do not constitute quality indicators.


2.  

    ```         
    How is the value-added measure (IVAL, indicateur de valeur ajoutée), calculated?[^sources2]
    ```

[^sources2]: 
  https://catalogue.depp.education.fr/index.php/catalog/274 
  Franck Evain (Insee), *Indicateurs de valeur ajoutée des lycées*:  https://www.insee.fr/fr/information/5008703?sommaire=5008710 

The **value-added indicator** measures the contribution of a high school to student outcomes, net of external factors.\
It is defined as the difference between the observed rate and the expected rate:

\[ \text{Value-added} = \text{Observed rate} - \text{Expected rate} \]

The expected rate is the average rate of high schools with students that have identical characteristics. Therefore the IVAL evaluates the contribution due to the establishment, given initial students’ profiles. A positive value-added indicates that on average the high school allows students to perform better than expected given their initial profile, while a negative value-added indicates underperformance.

\underline{Estimation of expected rates}

Expected rates are estimated using **multilevel logistic regression models**, which are appropriate for binary outcomes (success or failure, mention or no mention) and account for the hierarchical structure of the data (students nested within schools).

The models include:

**Individual-level variables** - Age\
- Gender\
- Academic achievement at entry into high school (e.g. results at the Diplôme National du Brevet)\
- Social background, measured using the Social Position Index (IPS)

**School-level variables** - Proportion of students with academic delay\
- Proportion of girls\
- Average IPS of students\
- Average prior academic achievement within the school

For each student, the model produces an expected probability of success, access, or obtaining a mention, assuming enrollment in a statistically average high school (i.e. excluding the specific school effect).\
The expected rate of a high school is then computed as the average of its students’ expected probabilities.

In the case of access to the baccalaureate, separate models are estimated for each transition (seconde→première, première→terminale, terminale→bac), and the expected access rate is obtained by multiplying the expected transition probabilities.

3.  

    ```         
    How do the IVAL and other indicators correlate with IPS (indicateurs de position sociale, a higher IPS indicates the higher average socio-economic status of parents at the school level) in Paris? Can you propose a way to represent graphically the correlations, for High schools in Paris (separately for public and private high schools)?
    ```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)

rm(list = ls())
```

```{r}
library(vroom)
library(here)
library(ggplot2)
library(dplyr)
library(readxl)
library(tidyr)
library(stringr)
library(janitor)
here::i_am("quality_education.Rproj")

# Import datasets
quality_indicators <- vroom(here("data", "fr-en-indicateurs-de-resultat-des-lycees-gt_v2.csv")) |>
  clean_names()
ips <- vroom(here("data", "fr-en-ips_lycees.csv")) |>
  clean_names()
path_ival <- here("data", "ival-2024-154664.xlsx")
ival_lgt <- read_excel(path_ival, sheet = "LGT", skip = 1) |>
  clean_names()
ival_lp  <- read_excel(path_ival, sheet = "LP",  skip = 1) |>
  clean_names()
```

```{r}
# Data cleaning and merging

# Convert logical columns (mostly NA) to numeric
quality_indicators <- quality_indicators |>
  mutate(across(where(is.logical), ~ as.numeric(.)))

# Build year from school-year label (e.g., "2023-2024" -> 2023)
ips <- ips |>
  mutate(annee = as.integer(str_extract(rentree_scolaire, "\\d{4}")))

# Keep Paris (dept 75)
quality_paris <- quality_indicators |>
  filter(code_departement == "75")

ips_paris <- ips |>
  filter(code_du_departement == "75")

# Pick the most recent common year
years_common <- intersect(sort(unique(quality_paris$annee)), sort(unique(ips_paris$annee)))
if (length(years_common) == 0) stop("No common year in Paris (75).")
year_use <- max(years_common, na.rm = TRUE)

quality_paris_y <- quality_paris |>
  filter(annee == year_use)

ips_paris_y <- ips_paris |>
  filter(annee == year_use)

# Harmonize sector labels
quality_paris_y <- quality_paris_y |>
  mutate(
    secteur = case_when(
      str_detect(tolower(secteur), "public") ~ "Public",
      str_detect(tolower(secteur), "priv") ~ "Private",
      TRUE ~ as.character(secteur)
    )
  )

ips_paris_y <- ips_paris_y |>
  mutate(
    secteur = case_when(
      str_detect(tolower(secteur), "public") ~ "Public",
      str_detect(tolower(secteur), "priv") ~ "Private",
      TRUE ~ as.character(secteur)
    )
  )

# Merge (UAI + year + sector)
df_paris <- quality_paris_y |>
  left_join(
    ips_paris_y |>
      select(
        uai, annee, secteur,
        matches("^ips_"),
        matches("^ecart_type_de_l_ips")
      ),
    by = c("uai", "annee", "secteur")
  )

# Quick join check
df_paris |>
  summarise(
    n_qi = n(),
    n_ips_match = sum(!is.na(ips_ensemble_gt_pro)),
    pct_ips_match = mean(!is.na(ips_ensemble_gt_pro)) * 100
  ) |>
  print()
```


## représentations graphiques
### analyses descriptives

```{r}
ggplot(df_paris,
       aes(x = ips_ensemble_gt_pro, fill = secteur)) +
  geom_density(alpha = 0.4) +
  labs(
    x = "IPS",
    title = "IPS distribution : Private vs Public"
  )

ggplot(df_paris,
       aes(x = ecart_type_de_l_ips_voie_gt, fill = secteur)) +
  geom_density(alpha = 0.4) +
  labs(
    x = "IPS Standard deviation",
    title = "Intern social heterogeneity distribution"
  )
```


```{r}
# nuages de points 
# 1. avec taux de réussite / IPS (+ lm)
ggplot(df_paris,
       aes(x = ips_ensemble_gt_pro,
           y = taux_de_reussite_toutes_series)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ secteur) +
  labs(
    x = "IPS",
    y = "Success rate",
    title = "IPS and Success rate"
  )

# 2. avec le taux d'accès
ggplot(df_paris,
       aes(x = ips_ensemble_gt_pro,
           y = taux_dacces_terminale_bac)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ secteur) +
  labs(
    x = "IPS",
    y = "Rate of access to the final year (Terminale)",
    title = "IPS and Rate of access to the final year"
  )

# 3. avec valeur ajoutée du taux de réussite
ggplot(df_paris,
       aes(x = ips_ensemble_gt_pro,
           y = valeur_ajoutee_du_taux_de_reussite_toutes_series,
           size = effectif_de_terminale)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ secteur) +
  labs(
    x = "IPS",
    y = "added value of the Success rate",
    size = "Number of students in final year (terminale)",
    title = "IPS, VA and highschool size"
  )


```

```{r}
# densité

ggplot(df_paris,
       aes(x = valeur_ajoutee_du_taux_de_reussite_toutes_series,
           fill = secteur)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Added value of the Success rate",
    title = "distribution of the added value of the Success rate: Private vs Public"
  )
```






